Advancing Sustainable Mobility: A GIS-Based Multi-Criteria Decision Analysis for Equitable Electric Vehicle Supply Equipment Deployment in Philadelphia

### 2025-03-09 code/MCDA-EVSE.Rmd
major update

Version 2.0 \| First Created Nov 17, 2023 \| Updated Jun 2, 2024

@@

evse <- read.csv(here("data", "raw", "existing_evse.csv"))
evse <- evse %>% 
  filter(City == "Philadelphia") %>% 
  dplyr::select(Latitude, Longitude, EV.Level2.EVSE.Num) %>% 
  filter(is.na(Latitude) == FALSE & is.na(Longitude) == FALSE) %>% 
  st_as_sf(., coords = c("Longitude", "Latitude"), crs = 4326) %>% 
  st_transform('ESRI:102728') %>% 
  mutate(num = ifelse(is.na(EV.Level2.EVSE.Num), 1, EV.Level2.EVSE.Num))

@@

ggplot()+
  geom_sf(data=fishnet, color="black", fill="white") +
  geom_sf(data=WSM, color="black", aes(fill=quantile)) + 
  scale_fill_manual(values = custom_palette) + 
  labs(title = "Weighted Sum Method Score", fill = "Bins") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        )
```

@@

quantiles <- classIntervals(TOPSIS_toplot$values, n = 6, style = "quantile")
TOPSIS_toplot$quantile <- cut(TOPSIS_toplot$values, breaks = quantiles$brks, include.lowest = TRUE)

ggplot()+
  geom_sf(data=fishnet, color="black", fill="white") +
  geom_sf(data=TOPSIS_toplot, color="black", aes(fill=quantile))+ 
  scale_fill_brewer(palette = "Blues") + 
  labs(title = "TOPSIS Score", fill = "Bins") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        )
```

@@

```{r visualize promethee}
quantiles <- classIntervals(PROMETHEE_toplot$net, n = 6, style = "quantile")
PROMETHEE_toplot$quantile <- cut(PROMETHEE_toplot$net, breaks = quantiles$brks, include.lowest = TRUE)

ggplot()+
  geom_sf(data=fishnet, color="black", fill="white") +
  geom_sf(data=PROMETHEE_toplot, color="black", aes(fill=quantile)) + 
  scale_fill_brewer(palette = "Blues") +
  labs(title = "PROMETHEE Score", fill = "Bins") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        )
```

@@

AHP_TOPSIS_toplot <- fishnet_nowater %>% 
  left_join(TOPSIS, by = "uniqueID")

@@

ggplot()+
  geom_sf(data=fishnet, color="black", fill="white") +
  geom_sf(data=AHP_TOPSIS_toplot, color="black", aes(fill=quantile)) + 
  scale_fill_manual(values = custom_palette) +
  labs(title = "AHP Score", fill = "Bins") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        )

@@

## Average Rank

@@

st_write(rank_compare, here("data", "derived", "average_rank.geojson"), driver = "GeoJSON")

@@

We took the average rank among all methods because we would like to find places in Philadelphia that are consistently ranked more suitable than other.

```{r visualize the average rank}

rank_toplot <- fishnet_nowater %>% 
  left_join(rank_compare %>% st_drop_geometry(), by = "uniqueID")
ggplot()+
  geom_sf(data=fishnet, color="black", fill="white") +
  geom_sf(data=rank_toplot, color="black", aes(fill=avg_rank)) +
  labs(title = "Average Rank", fill = "Rank") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        )

## Rank by Neighborhood

We would like to see if some neighborhoods are more suitable than others overall.

```{r add neighborhood}

neighborhood <- st_read(here("data", "raw", "PhillyPlanning_Neighborhoods", "PhillyPlanning_Neighborhoods.shp")) %>% st_transform('ESRI:102728')

@@

```{r neighborhood plot}

final %>% 
  st_drop_geometry() %>% 
  group_by(NAME) %>% 
  summarize(overall_rank = mean(avg_rank)) %>% 
  left_join(neighborhood, .,by = "NAME") %>% 
  ggplot()+
  geom_sf(color="black", aes(fill=overall_rank)) +
  scale_fill_gradient(low = "red", high = "white") + # Changed this line
  labs(title = "Neighborhood Rank", fill = "Rank") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        )

@@

## Compare MCDA Methods

```{r compare difference distribution, fig.height=6, fig.width=10}

# if negative, it means that wsm over-rank topsis or promethee
# if positive, it means that wsm under-rank topsis or promethee
compare<- final %>% 
  mutate(wsm_topsis = wsm_rank - topsis_rank, # compare the rank between wsm and topsis
         wsm_promethee = wsm_rank - promethee_rank, # compare the rank between wsm and promethee
         promethee_topsis = promethee_rank - topsis_rank) %>% # compare the rank between promethee and topsis
  mutate(compare_weights = topsis_rank - ahptopsis_rank)

custom_palette1 <- c("#B2BF95", "#80A676", "#88BFBF", "#7EA1BF")

fig1 <- compare %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(uniqueID, wsm_topsis, wsm_promethee, promethee_topsis, compare_weights, NAME)) %>%
  pivot_longer(cols = c("wsm_topsis", "wsm_promethee", "promethee_topsis", "compare_weights"),  
               names_to = "variable",        
               values_to = "value") %>% 
  ggplot(aes(x = value)) +
  geom_histogram(binwidth = 10, aes(fill = variable), alpha = 1) +
  labs(title = "Distribution of Differences in Rank Between Different Methods",
       x = "Difference",
       y = "Frequency") +
  facet_wrap(~ variable, labeller= labeller(variable = c(
    `promethee_topsis` = "PROMETHEE  vs. TOPSIS ",
    `wsm_topsis` = "WSM  vs. TOPSIS",
    `wsm_promethee` = "WSM  vs. PROMETHEE",
    `compare_weights` = "AHP and Directly Assign Weights"))) +
  scale_fill_manual(values = custom_palette1) + 
  theme(#axis.text.x=element_blank(),
        #axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8),
        legend.position = "none",  # Turn off the legend,
        strip.text = element_text(size = 12)
        )

fig1 
#ggsave("graph.png", fig1, width = 10, height = 6, dpi = 300)
```

@@

my_sort <- c("promethee_topsis", "wsm_topsis", "wsm_promethee", "compare_weights")

# Your data manipulation and ggplot code
plot <- compare %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(uniqueID, wsm_topsis, wsm_promethee, promethee_topsis, compare_weights, NAME)) %>%
  pivot_longer(cols = c("wsm_topsis", "wsm_promethee", "promethee_topsis", "compare_weights"),  
               names_to = "variable",        
               values_to = "value") %>% 
  ggplot(aes(x = as.numeric(factor(variable, levels = my_sort)))) +
  geom_half_violin(aes(x = as.numeric(factor(variable, levels = my_sort)) + 0.1,
                       y = value, fill = factor(variable, levels = my_sort)),
                   side = 'r', scale = "width", cex = 0.8, lwd=0.3) +  
  geom_boxplot(aes(x = as.numeric(factor(variable, levels = my_sort)) + 0.1,
                   y = value, fill = factor(variable, levels = my_sort)),
               outlier.colour = "black", width = 0.1, cex = 0.8, lwd=0.3, outlier.size=0.6) +  
  geom_jitter(aes(x = as.numeric(factor(variable, levels = my_sort)) - 0.1,
                  y = value, color = factor(variable, levels = my_sort)),
              width = 0.1, size = 0.2, stroke = 0.8) +
  scale_fill_manual(values = custom_palette1) + 
  scale_color_manual(values = custom_palette1) +  
  labs(title = "Distribution of Differences in Rank Between Different Methods",
       x = "Method",
       y = "Difference") +
  theme(axis.text.x=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = "none",  
        )

# Print the plot
print(plot)

#ggsave("boxplot.png", plot, width = 10, height = 6, dpi = 300)

@@

@@

custom_palette2 <- c("#80A676", "white", "#7EA1BF")

fig2 <- compare %>% 
  dplyr::select(c(uniqueID, wsm_topsis, wsm_promethee, promethee_topsis, compare_weights, NAME)) %>%
  pivot_longer(cols = c("wsm_topsis", "wsm_promethee", "promethee_topsis", "compare_weights"),  
               names_to = "variable",       
               values_to = "value") %>% 
  filter(variable == "promethee_topsis" | variable == "wsm_topsis") %>% 
  ggplot() +
  geom_sf(aes(fill = value), color = "grey") +
  labs(title = "Ranking Difference between TOPSIS and Other Approaches") +
  facet_wrap(~ variable, labeller= labeller(variable = c(
    `promethee_topsis` = "PROMETHEE  vs. TOPSIS ",
    `wsm_topsis` = "WSM  vs. TOPSIS",
    `wsm_promethee` = "WSM  vs. PROMETHEE",
    `compare_weights` = "AHP and Directly Assign Weights"))) +
  scale_fill_gradientn(colors = custom_palette2, 
                       name = "Rank Difference", 
                       na.value = "grey50") + 
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank()    )

fig2
#ggsave("plot.png", fig2, width = 10, height = 6, dpi = 300)

@@

@@

```{r}
## plot top 20 loadings
top_k <- 5

## get pc1 and pc2
pc1 <- data.frame(loading = pca_fishnet$rotation[,1],
                  var = rownames(pca_fishnet$rotation),
                  pc = "PC1") 
pc2 <- data.frame(loading = pca_fishnet$rotation[,2],
                  var = rownames(pca_fishnet$rotation),
                  pc = "PC2")

# get top_k loadings of pc1 and pc2
pc1_top <- pc1 %>% arrange(-loading) %>% slice(1:top_k)
pc2_top <- pc2 %>% arrange(-loading) %>% slice(1:top_k)

rbind(pc1_top, pc2_top) %>%
  ggplot(aes(x = reorder(var, -loading), y = loading)) +
  geom_point() +
  ggtitle("Top loadings") +
  xlab("Gene") +
  facet_wrap(~pc, nrow = 1, scales = "free_x") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = -45, hjust = 0, vjust = 1))
```

@@

@@

fishnet_clustered %>%
  st_drop_geometry() %>%
  dplyr::select(Cluster, driving_pop, evse.nn, elec_cnt, parking.nn) %>%
  mutate(
         evse.nn = scale_values(evse.nn),
         driving_pop = scale_values(driving_pop),
         elec_cnt = scale_values(elec_cnt),
         parking.nn= scale_values(parking.nn)
         ) %>% 
  pivot_longer(cols = -Cluster, names_to = "Variable", values_to = "Scaled_Value") %>% 
  ggplot(., aes(x = as.factor(Cluster), y = Scaled_Value, fill = as.factor(Cluster))) +
  geom_violin(trim = FALSE, size = 0.4)+
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set3") +
  facet_wrap(~Variable, scales = "free_y") +  # Facet by variable
  labs(
    title = "Distribution of Scaled Criteria by Cluster",
    x = "Cluster",
    y = "Scaled Value (0-1)",
    fill = "Cluster"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(size = 10),
    legend.position = "none",  # Remove legend since colors already represent clusters
    strip.text = element_text(size = 11, face = "bold")  # Styling facet labels
  )
```

@@

```{r original percentile}

# Get the original rank and compute percentile rank 
# then join it with the result from local MCDA
# to compare whether the percentile rank is higher or lower than the global MCDA
final_percentile <- final %>% 
  dplyr::select(NAME, uniqueID, avg_rank, topsis_rank, wsm_rank) %>%
  mutate(
    avg_percentile = percent_rank(avg_rank) * 100,
    wsm_percentile = percent_rank(wsm_rank) * 100,
    topsis_percentile = percent_rank(topsis_rank) * 100)

## Try Local WSM

```{r}
# weights_c1 <- c(0.12, 0.14, 0.12, 0.16, 0.16, 0.08, 0.06, 0.06, 0.10)
WSM_c1 <- cluster1 %>% 
  mutate(scl_density = scale_values(density),
         scl_acc = scale_values(TSFCA),
         scl_drivingpop = scale_values(driving_pop),
         scl_evse = scale_values(evse.nn),
         scl_elec = scale_values(elec_cnt),
         scl_parking = scale_values(parking.nn),
         scl_policefire = scale_values(policefire.nn),
         scl_elect_use = scale_values(elect_use)
         ) %>% 
  mutate(scl_evse_re = 0 - scl_evse + 1,
         scl_acc_re = 0 - scl_acc + 1, 
         scl_parking_re = 0 - scl_parking + 1,
         scl_policefire_re = 0 - scl_policefire + 1) %>% 
  mutate(score = 0.12*scl_evse_re + 0.14*scl_density + 0.12*scl_drivingpop + 0.16*scl_acc_re + 0.16*scl_elec + 0.08*scl_parking_re + 0.06*scl_policefire_re + 0.06*zone_score + 0.10*scl_elect_use) %>% 
  dplyr::select(c(uniqueID, score)) %>% 
  mutate(wsm_rank_c1 = min_rank(desc(score))) %>% 
  mutate(wsm_percentile_c1 = percent_rank(wsm_rank_c1) * 100)

#weights_c2 <- c(0.14, 0.08, 0.08, 0.10, 0.12, 0.12, 0.06, 0.10, 0.20)
WSM_c2 <- cluster2 %>% 
  mutate(scl_density = scale_values(density),
         scl_acc = scale_values(TSFCA),
         scl_drivingpop = scale_values(driving_pop),
         scl_evse = scale_values(evse.nn),
         scl_elec = scale_values(elec_cnt),
         scl_parking = scale_values(parking.nn),
         scl_policefire = scale_values(policefire.nn),
         scl_elect_use = scale_values(elect_use)
         ) %>% # now inverse some scales
  mutate(scl_evse_re = 0 - scl_evse + 1,
         scl_acc_re = 0 - scl_acc + 1, 
         scl_parking_re = 0 - scl_parking + 1,
         scl_policefire_re = 0 - scl_policefire + 1) %>% 
  mutate(score = 0.14*scl_evse_re + 0.08*scl_density + 0.08*scl_drivingpop + 0.10*scl_acc_re + 0.12*scl_elec + 0.12*scl_parking_re + 0.06*scl_policefire_re + 0.10*zone_score + 0.20*scl_elect_use) %>% 
  dplyr::select(c(uniqueID, score)) %>%
  mutate(wsm_rank_c2 = min_rank(desc(score))) %>%
  mutate(wsm_percentile_c2 = percent_rank(wsm_rank_c2) * 100)

>>>>

Version 5.0 \| First Created Nov 17, 2023 \| Updated Mar 9, 2025

@@

evse <- read.csv(here::here("data", "raw", "existing_evse.csv"))
evse <- evse %>% 
  filter(City == "Philadelphia") %>% 
  dplyr::select(Latitude, Longitude, EV.Level2.EVSE.Num) %>% 
  filter(is.na(Latitude) == FALSE & is.na(Longitude) == FALSE) %>% 
  st_as_sf(., coords = c("Longitude", "Latitude"), crs = 4326) %>% 
  st_transform('ESRI:102728') %>% 
  mutate(num = ifelse(is.na(EV.Level2.EVSE.Num), 1, EV.Level2.EVSE.Num))

@@

WSM_plot <- ggplot()+
  geom_sf(data=fishnet, color=NA, fill="white") +
  geom_sf(data=WSM, color = NA, aes(fill=quantile)) + 
  scale_fill_manual(values = custom_palette) + 
  labs(title = "Weighted Sum Method Score", fill = "Bins") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        )
```

@@

quantiles <- classIntervals(TOPSIS_toplot$values, n = 5, style = "quantile")
TOPSIS_toplot$quantile <- cut(TOPSIS_toplot$values, breaks = quantiles$brks, include.lowest = TRUE)

TOPSIS_plot <- ggplot()+
  geom_sf(data=fishnet, color=NA, fill="white") +
  geom_sf(data=TOPSIS_toplot, color=NA, aes(fill=quantile))+ 
  scale_fill_manual(values = custom_palette) + 
  labs(title = "TOPSIS Score", fill = "Bins") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        )

@@

```{r visualize promethee}
quantiles <- classIntervals(PROMETHEE_toplot$net, n = 5, style = "quantile")
PROMETHEE_toplot$quantile <- cut(PROMETHEE_toplot$net, breaks = quantiles$brks, include.lowest = TRUE)

PROMETHEE_plot <- ggplot()+
  geom_sf(data=fishnet, color=NA, fill="white") +
  geom_sf(data=PROMETHEE_toplot, color=NA, aes(fill=quantile)) + 
  scale_fill_manual(values = custom_palette) + 
  labs(title = "PROMETHEE Score", fill = "Bins") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        )

@@

AHP_TOPSIS_toplot <- fishnet_nowater %>% 
  left_join(AHP_TOPSIS, by = "uniqueID")

@@

AHP_TOPSIS_plot <- ggplot()+
  geom_sf(data=fishnet, color=NA, fill="white") +
  geom_sf(data=AHP_TOPSIS_toplot, color=NA, aes(fill=quantile)) + 
  scale_fill_manual(values = custom_palette) +
  labs(title = "AHP Score", fill = "Bins") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)
        )

@@

## Mapping Spatial Consistency

```{r fig.height=12, fig.width=12}

library(ggpubr)
ggarrange(WSM_plot, TOPSIS_plot, PROMETHEE_plot, AHP_TOPSIS_plot, ncol = 2, nrow = 2)

## Measure Rank Alignment

@@

```{r spearman rank correlation}

options(scipen=999)
# Define a function to compute correlation and p-value
compute_spearman <- function(x, y) {
  cor_test <- cor.test(x, y, method = "spearman")
  return(c(correlation = cor_test$estimate, p_value = cor_test$p.value))
}

# Compute correlations and p-values for all comparisons
global_cor_results <- data.frame(
  Comparison = c("AHP vs TOPSIS", "AHP vs WSM", "AHP vs PROMETHEE",
                 "TOPSIS vs WSM", "TOPSIS vs PROMETHEE", "WSM vs PROMETHEE"),
  Correlation = c(
    compute_spearman(rank_compare$ahptopsis_rank, rank_compare$topsis_rank)[1],
    compute_spearman(rank_compare$ahptopsis_rank, rank_compare$wsm_rank)[1],
    compute_spearman(rank_compare$ahptopsis_rank, rank_compare$promethee_rank)[1],
    compute_spearman(rank_compare$topsis_rank, rank_compare$wsm_rank)[1],
    compute_spearman(rank_compare$topsis_rank, rank_compare$promethee_rank)[1],
    compute_spearman(rank_compare$wsm_rank, rank_compare$promethee_rank)[1]
  ),
  P_Value = c(
    compute_spearman(rank_compare$ahptopsis_rank, rank_compare$topsis_rank)[2],
    compute_spearman(rank_compare$ahptopsis_rank, rank_compare$wsm_rank)[2],
    compute_spearman(rank_compare$ahptopsis_rank, rank_compare$promethee_rank)[2],
    compute_spearman(rank_compare$topsis_rank, rank_compare$wsm_rank)[2],
    compute_spearman(rank_compare$topsis_rank, rank_compare$promethee_rank)[2],
    compute_spearman(rank_compare$wsm_rank, rank_compare$promethee_rank)[2]
  )
)
cor.test(rank_compare$ahptopsis_rank, rank_compare$topsis_rank, method = "spearman")

@@

```{r neighborhood?}

# rank consistency by neighborhood? 
neighborhood <- st_read(here("data", "raw", "PhillyPlanning_Neighborhoods", "PhillyPlanning_Neighborhoods.shp")) %>% st_transform('ESRI:102728')

@@

1. Difference between WSM and TOPSIS
2. Difference between WSM and PROMETHEE
3. Difference between WSM and AHP
4. Difference between TOPSIS and PROMETHEE
**5. Difference between TOPSIS and AHP**
6. Difference between PROMETHEE and AHP

# if negative, it means that the first one over-rank the second one
# if positive, it means that the first one under-rank the second one
compare<- final %>% 
  mutate(wsm_topsis = wsm_rank - topsis_rank, # compare the rank between wsm and topsis
         wsm_promethee = wsm_rank - promethee_rank, # compare the rank between wsm and promethee
         wsm_ahp = wsm_rank - ahptopsis_rank, # compare the rank between wsm and ahp
         promethee_topsis = promethee_rank - topsis_rank, # compare the rank between topsis and promethee
         topsis_ahp = topsis_rank - ahptopsis_rank, # compare the rank between topsis and ahp
         promethee_ahp = promethee_rank - ahptopsis_rank)  # compare the rank between promethee and topsis

@@

```{r compare difference distribution 2}

custom_palette1 <- c("#B2BF95", "#80A676", "#88BFBF", "#7EA1BF", "grey", "pink")
my_sort <- c("promethee_topsis", "topsis_ahp", "wsm_topsis", "wsm_promethee", "wsm_ahp", "promethee_ahp")

temp <- compare %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(uniqueID, promethee_topsis, topsis_ahp, wsm_topsis, wsm_promethee, wsm_ahp, promethee_ahp, NAME)) %>%
  pivot_longer(cols = c("promethee_topsis", "topsis_ahp", "wsm_topsis", "wsm_promethee", "wsm_ahp", "promethee_ahp"),  
               names_to = "variable",        
               values_to = "value") %>% 
  ggplot(aes(x = as.numeric(factor(variable, levels = my_sort)))) +
  geom_half_violin(aes(x = as.numeric(factor(variable, levels = my_sort)) + 0.1,
                       y = value, fill = factor(variable, levels = my_sort)),
                   side = 'r', scale = "width", cex = 0.8, lwd=0.3) +  
  geom_boxplot(aes(x = as.numeric(factor(variable, levels = my_sort)) + 0.1,
                   y = value, fill = factor(variable, levels = my_sort)),
               outlier.colour = "black", width = 0.1, cex = 0.8, lwd=0.3, outlier.size=0.6) +  
  geom_jitter(aes(x = as.numeric(factor(variable, levels = my_sort)) - 0.1,
                  y = value, color = factor(variable, levels = my_sort)),
              width = 0.1, size = 0.2, stroke = 0.8) +
  scale_fill_manual(values = custom_palette1) + 
  scale_color_manual(values = custom_palette1) +  
  labs(title = "Distribution of Differences in Rank Between Different Methods",
       x = "Method",
       y = "Difference") +
  theme_bw()
  theme(axis.text.x=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_blank(),
        )

# my_sort <- c("topsis_promethee", "topsis_ahp", "wsm_topsis", "wsm_promethee", "wsm_ahp", "promethee_ahp")
ggsave("/Users/emzhou/Desktop/Paper/svg/violin.svg", temp, width = 10, height = 7)

```{r}
# map distribution of the criteria
final %>% 
  st_drop_geometry() %>%
  dplyr::select(c(density, TSFCA, driving_pop, evse.nn, elec_cnt, parking.nn, policefire.nn, elect_use, zone_score)) %>%
  pivot_longer(cols = c(density, TSFCA, driving_pop, evse.nn, elec_cnt, parking.nn, policefire.nn, elect_use, zone_score),  
               names_to = "variable",        
               values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_density(aes(fill = variable), alpha = 0.5) +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Distribution of Criteria",
       x = "Value",
       y = "Density")

summary(final$elect_use)
```

```{r }

final_sankey <- final %>%
  mutate(topsis_percentile = percent_rank(topsis_rank),
         wsm_percentile = percent_rank(wsm_rank),
         promethee_percentile = percent_rank(promethee_rank),
         ahp_percentile = percent_rank(ahptopsis_rank)) %>%
  mutate(
    topsis_cat = cut(topsis_percentile, breaks = c(-1, 0.2, 0.4, 0.6, 0.8, 1.1), labels = c("0-20%", "20-40%", "40-60%", "60-80%", "80-100%")),
    wsm_cat = cut(wsm_percentile, breaks = c(-1, 0.2, 0.4, 0.6, 0.8, 1.1), labels = c("0-20%", "20-40%", "40-60%", "60-80%", "80-100%")),
    promethee_cat = cut(promethee_percentile, breaks = c(-1, 0.2, 0.4, 0.6, 0.8, 1.1), labels = c("0-20%", "20-40%", "40-60%", "60-80%", "80-100%")),
    ahp_cat = cut(ahp_percentile, breaks = c(-1, 0.2, 0.4, 0.6, 0.8, 1.1), labels = c("0-20%", "20-40%", "40-60%", "60-80%", "80-100%"))
  )

final_sankey %>% 
  dplyr::count(topsis_cat, wsm_cat, promethee_cat, ahp_cat) %>%
  ggplot(., aes(axis1 = topsis_cat, axis2 = ahp_cat, axis3 = promethee_cat, axis4 = wsm_cat, y = n)) +
  scale_x_discrete(limits = c("TOPSIS", "AHP", "PROMETHEE", "WSM"), 
                   expand = c(.05, .5)) +
  geom_alluvium(aes(fill = topsis_cat), alpha = 0.8) +  # Fill based on initial ranking
  geom_stratum(color = "grey") + 
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3) + 
  scale_fill_manual(values = custom_palette) +
  theme_bw() +
  labs(title = "Rank Transitions Between Global and Local TOPSIS Methods",
       x = "Ranking Method", y = "Grid Count")

@@

# percent of identical positions in different decision-making methods
final_sankey %>% 
  dplyr::select(topsis_cat, wsm_cat, promethee_cat, ahp_cat) %>%
  #filter(promethee_cat == wsm_cat) %>%
  #filter(ahp_cat == wsm_cat) %>%
  # filter(promethee_cat == ahp_cat) %>%
  #filter(topsis_cat == ahp_cat) %>%
  #filter(topsis_cat == wsm_cat) %>%
  #filter(topsis_cat == promethee_cat) %>%
  filter(wsm_cat == ahp_cat & ahp_cat == promethee_cat)

@@

Pct of identical positions between TOPSIS and WSM: 772 / 1212 = 63.7%
Pct of identical positions between TOPSIS and PROMETHEE: 750 / 1212 = 61.9%
Pct of identical positions between WSM and PROMETHEE: 848 / 1212 = 70.0%
Pct of identical positions between TOPSIS and AHP-Topsis: 805 / 1212 = 66.4%
Pct of identical positions between WSM and AHP-Topsis: 696 / 1212 = 57.4%
Pct of identical positions between PROMETHEE and AHP-Topsis: 654 / 1212 = 54.0%
Pct of identical positions between TOPSIS, AHP-TOPSIS, and WSM: 568 / 1212 = 46.9%
Pct of identical positions between TOPSIS, AHP-TOPSIS, and PROMETHEE: 548 / 121 = 45.2%
Pct of identical positions between WSM, AHP-TOPSIS, and PROMETHEE: 545 / 1212 = 45.0%
Pct of identical positions between all four methods: 472 / 1212 = 38.9%

## Visualize Spatial Rank Variability

@@

custom_palette2 <- c("#e8623f", "white", "#57748d")

compare %>% 
  dplyr::select(c(uniqueID, promethee_topsis, topsis_ahp, wsm_topsis, wsm_promethee, wsm_ahp, promethee_ahp, NAME)) %>%
  pivot_longer(cols = c("promethee_topsis", "wsm_topsis",  "wsm_ahp", "promethee_ahp"),  
               names_to = "variable",       
               values_to = "value") %>% 
  # filter(variable == "promethee_topsis" | variable == "wsm_topsis") %>% 
  ggplot() +
  geom_sf(aes(fill = value), color = "grey") +
  labs(title = "Ranking Difference between TOPSIS and Other Approaches") +
  facet_wrap(~ variable, labeller= labeller(variable = c(
    `promethee_topsis` = "PROMETHEE vs. TOPSIS",
    `topsis_ahp` = "TOPSIS  vs. AHP",
    `wsm_topsis` = "WSM  vs. TOPSIS",
    `wsm_promethee` = "WSM  vs. PROMETHEE",
    `wsm_ahp` = "WSM  vs. AHP",
    `promethee_ahp` = "PROMETHEE  vs. AHP"))) +
  scale_fill_gradientn(colors = custom_palette2, 
                       name = "Rank Difference", 
                       na.value = "grey50") + 
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank()    )

st_write(compare, here::here("data", "derived", "compare.geojson"), driver = "GeoJSON")

@@

## Cluster Determination

@@

@@

## Cluster Characteristics and Local Weights

@@

fishnet_clustered %>%
  st_drop_geometry() %>%
  dplyr::select(Cluster, driving_pop, evse.nn, elec_cnt, parking.nn) %>%
  mutate(
    evse.nn = scale_values(evse.nn),
    driving_pop = scale_values(driving_pop),
    elec_cnt = scale_values(elec_cnt),
    parking.nn = scale_values(parking.nn),
  ) %>%
  pivot_longer(cols = -Cluster, names_to = "Variable", values_to = "Scaled_Value") %>%
  ggplot(aes(x = Scaled_Value, fill = as.factor(Cluster))) +
  geom_density(alpha = 0.6) +
  scale_fill_brewer(palette = "Set3") +
  facet_wrap(~Variable, scales = "free") +
  labs(
    title = "Density Distribution of Scaled Criteria by Cluster",
    x = "Scaled Value",
    y = "Density",
    fill = "Cluster"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(size = 10),
    legend.position = "bottom",
    strip.text = element_text(size = 11, face = "bold")
  )

@@

## Local MCDA Using TOPSIS
